---
title: "Introduction to R"
author: "Sarah Meister"
date: '2015-11-02'
output: pdf_document
fontsize: 8pt
header-includes: \input{../slideOptions.tex}
layout: page
classoption: xcolor=dvipsnames
sidebar: no
slide-level: 1
tag:
- Lessons
- Slides
- IntroR
categories:
- Lessons
- IntroR
---
R is an object-oriented programing language, meaning everything we incounter in R is an object. 

R has several types of objects:

R has a 5 basic classes of objects:

* Character
* Numeric
* Integer
* Complex
* Logical

R has 5 different types of objects:

* Vectors
* Lists
* Matrices
* Factors
* Dataframes

Thus you could have a logical vector, a character dataframe, a numeric matrix, etc. 

# Vectors #

The most basic object is a vector. A vector is a sequence of objects of the same class. We can make vectors using the `c()` function.

```{r , collapse=TRUE}
a <- c(0.4, 1.2)         # numeric
b <- c(TRUE, FALSE)      # logical
c <- c("a","b","c")      # character
d <- c(1L, 2L)           # integer
e <- 1:10                # integer
f <- c(1+0i, 2+4i)       # complex
```

Empty vectors can be made using the `vector()` function

```{r , collapse=TRUE}
x <- vector("numeric", length=10)
x
```

# Lists #

A list is a special type of vector that can contain objects of different classes. For instance, we can have a list with a numeric dataframe as one object within the list, and a string vector as another. 

```{r , collapse=TRUE}
x <- list(1,"a", TRUE, 1+4i)
x
```

The elements of a list are seperated by double brackets. 
 
# Matrices #

Matrices are vectors with a dimension attribute. The dim attribute is an integer vector of length = 2 (nrow, ncol). Matrix multiplication is done with the %*% operator

```{r , collapse=TRUE}
x <- matrix(1:6, nrow=2, ncol=3)

dim(x)

attributes(x)
```
 
Matrices are constructed column-wise, so entries start at the upper left corner and run down the columns. 

Matrices can be created directly from vectors by adding a dimension value

```{r, collapse=TRUE}
x <- 1:10
x
dim(x) <- c(2,5)
x
```

We can also create matrices by column-binding or row-binding with the functions `cbind()` and `rbind()`

```{r, collapse=TRUE}
x <- 1:3
y <- 10:12
cbind(x,y)
rbind (x, y)
```

# Factors #

Factors are used to represent catagorical data. They can be ordered or unordered.

* They are like an integer vector where each integer has a label. 
* Factors are more descriptive than integers e.g. "Male" and "Female" vs. 1 and 2
* Factors are treated specially by modelling functions like `lm()` and `glm()`

```{r, collapse=TRUE}
x <- factor(c("yes", "yes", "no", "no", "yes"))
x
table (x)
unclass(x)
```

# Dataframe #

Dataframes are used in R to store tabular data. They are represented in R as a special type of list where every element in the list has the same length. Each element of the list can be thought of as a **column** and each length of the list can be thought of as a **row**. Because a dataframe is a type of list, dataframes can store different classes of objects in each column (e.g. numeric, character, logical, etc.).

Depending on the source of the data, dataframes can be made using the `read.table()`, `read.csv` or `data.frame()` functions. 

```{r, collapse=TRUE}
x <- data.frame(foo = 1:4, bar = c(T,T,F,F))
x
nrow(x)
ncol(x)
```


# Attributes #

R objects can have attributes, such as:

* names
* classes
* lengths
* dimensions (matrices/dataframes)
* user defined attributes, such as metadata
* dataframes have a special attribute called `row.names`

Attributes can be modified with the `attributes()` function, among others.

For example, modifiying the `names` of a vector can make your code more readable.
```{r, collapse=TRUE}
x <- 1:3
names(x)

names(x) <- c("alpha", "beta", "gamma")
x
names(x)
```

# Looking for help in R #

* Look up the documentation for a function: `help("function")`

OR

?function

* Look up documentation for a package: `help(package="packagename")`

# Packages #

R comes with a few default packages, however there are thousands of R packages that extend R's capabilities (such as `ggplot2`, our next lesson)

* To see what packages are loaded: `search()`

* To view available packages: `library()`

* To load a package: `library(packagename)`

* Install new package: `install.packages("packagename")` 

# Working Directory #

R knows the directory it was started in, and refers to this as the "working directory". This is where R will search for any files you are imputting, and where it will place any files you output.  

To look at the current working directory: `getwd()` 

To change the working directory: ` setwd("directory_path")`

To make a new folder in R: `dir.create(path)`

# Imputting Tabular Data #

Most data you will work with comes in tabular form (such as an Excel spreadsheet). The two most commonly-used functions used to imput tabular data into R are `read.table()` and `read.csv()`. 
```{r, eval=FALSE, collapse=TRUE}
data <- read.table(file_name.txt, header=FALSE, sep="\t")

# \t is the regular expression for the tab key. This means our table is tab-delimited (tab-separated)

data2 <- read.csv(file_name.csv, header=TRUE)

#.csv is a comma-separated table 
```

# Outputting Tabular Data #

Often you'll want to save R data into a tabular form (e.g. to view in Excel). This is where you use `write.table()`
```{r, eval=FALSE, collapse=TRUE}
write.table(data, file= "myfile.csv", sep=",", row.names=TRUE, col.names=FALSE)
```

# Subsetting Data #

Using `[]` in R allows you to subset certain parts of a vector

```{r, collapse=TRUE}
x <- c(TRUE, FALSE, "alpha", "beta", "gamma", 1, 2)

x[c(1, 4, 5)]

```

This also works for matrices

```{r, collapse=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
a[1:2, ]
a[c(T, F, T), c("B", "A")]
a[0, -2]
```

and dataframes

```{r, collapse=TRUE}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])

df[ ,c(1, 3)]    #gives the columns that are in position 1 and 3
```

You can also easily subset a column in a dataframe using a $
```{r, collapse=TRUE}
df <- data.frame(x = 1:10, y = 11:20, z = letters[1:10])

df$z
```

# Logical and Relative Operators #

Arthmetic Operators

Operator | Description
-------- | -----------
+        | addition
-        | subtraction
*        | multiplication
/        | division
^ or **  | exponentiation
x %% y   | modulus (5 %% 2 is 1)
x %/% y    | integer division (5%/% is 2)
x %*% y | matrix multiplication

Logical Operators

Operator  | Description
--------  | -----------
<         | less than
<=        | less than or equal to
>         | greater than
>=        | greater than or equal to
==        | exactly equal to
!=        | not equal to
!x        | not x
x &#124; y    | x OR y (returns a vector)
x &#124;&#124; y    | x OR y (examines only 1st element of vector)
x & y     | x AND y (returns a vector)
x && y    | x AND y (examines only 1st element of vector)
isTRUE(x) | test if X is TRUE

# subset function #

The `subset()` function in R will return a subset of an argument that meet a particular condition. 

```{r, collapse=TRUE}
numvec = c(2,5,8,9,0,6,7,8,4,5,7,11)

subset(numvec, numvec < 9 & numvec > 4)

```

# if-else expressions #

```{r, eval=FALSE, collapse=TRUE}
if (condition) {
  ##do this
} else {
  ##do that
}

## to add more conditions

if (condition1) {
  ## do this
} else if (condition2) {
  ## do something different
} else {
  ## do something else
}
```

# if-else expressions #

```{r, collapse=TRUE, warning=FALSE}
if (x > 3) {
  y <- 10
} else {
  y <- 0
}


### OR


y <- if(x > 3) {
  10
} else {
  0
}
```

# for loop #
```{r, collapse=TRUE}
for (i in 1:10) {
  print (i)
}


### nested for loops

x <- matrix(1:6, 2, 3)

for (i in seq_len(nrow(x))) {
  for (j in seq_len(ncol(x))) {
    print (x[i, j])
  }
}
```

# Writing functions #

functions are created using the `function()` directive and are stored as R objects of the class "function". Functions can be passed as arguments to other functions and they can be nested (functions inside functions)

```{r, eval=FALSE, collapse=TRUE}
f <- function(argument) {
  ## do something here 
}
```

```{r, collapse=TRUE}
above10 <- function(x) {
  use <- x > 10
  x[use]
}

above <- function(x, n) {
  use <- x > n
  x[use]
}

columnmean <- function (y, remove.NA = TRUE) {
  nc <- ncol(y)
  means <- numeric(nc)
  for (i in 1:nc) {
    means[i] <- mean(y[,i], na.rm = remove.NA)
  }
  means
}
```
# Loop Functions #

R has built loop functions, which often have the word "apply" in them.

- `lapply`: loop over a list and evaluate a function on each element (useful with the `split` function)
- `sapply`: same as `lapply` but it tries to simplify the result
- `apply`: apply a function over the margins of an array 
- `tapply`: apply a function over subsets of a vector
- `mapply`: a multivariate version of lapply 

# lapply #

`lapply` loops internally with C code, making the looping faster

`lapply` takes two to three arguments:

1. a list `x` (or another object that is coerced to a list)
2. a function
3. other arguments as necessary

```{r, collapse=TRUE}
x <- list(a=1:5, b=rnorm(10), c=rnorm(20,1), d=rnorm(100,5))
lapply(x, mean)
```
`lapply` will always return a list

`lapply` and others make heavy use of anonymous functions, which are functions without names.
```{r, collapse=TRUE}
x <- list(a=matrix(1:4,2,2), b=matrix(1:6,3,2))
x

lapply(x, function(ele) ele[ ,1]) 
## create function with the argument ele and then define that argument.
```

# sapply #

`sapply` will try to simplify the result of lapply if possible

- if the result is a list where every element is length 1, than a vector is returned 
- if the result is a list where every element is a vector of the same length, a matrix is returned
- otherwise a list is returned

```{r, collapse=TRUE}
x <- list(a=1:5, b=rnorm(10), c=rnorm(20,1), d=rnorm(100,5))
sapply(x, mean)
```

# apply #

`apply` is used to evalute a function over the margins of an array. It is not faster than a loop, but it works in one line. `1` means do the function to the rows. `2` means do the function to the columns

```{r, collapse=TRUE}
x <- matrix (rnorm(200), 20, 10)
apply(x, 2, mean) # keeping number of columns and collapsing rows. This gives a vector of the means of columns. 

apply(x, 1, sum) # this calculates the sum of all rows

y <- matrix(rnorm(200), 20, 10)

apply(y, 1, quantile, probs = c(0.25, 0.75))

## with an array (stacks of matrices: multidimensional)

a <- array(rnorm(2*2*10), c(2,2,10)) #create an array that looks like a bunch of 2 by 2 matrices and take the mean of those
apply(a, c(1,2), mean) #this keeps the 1st and 2nd dimension

```

# mapply #

`mapply` is a multivariate apply (like `lapply`) that applies a function over a set of arguments. 

```{r, collapse=TRUE}
mapply(rep, 1:4, 4:1)

## is the same as

list(rep(1, 4), rep(2,3), rep(3,2), rep(4,1))
```

# Instant Vectorization #

```{r, collapse=TRUE}
noise<- function(n,mean, sd) {
  rnorm(n, mean, sd)
}

noise(5, 1, 2)

mapply(noise, 1:5, 1:5, 2)

## which is the same as

list(noise(1,1,2), noise(2,2,2), noise(3,3,2), noise(4,4,2), noise(5,5,2))
```

# tapply #

`tapply` is used to apply a function over subsets of a vector. 

```{r, collapse=TRUE}
x <- c(rnorm(10), runif(10), rnorm(10,1))
f<- gl(3,10) #factor variable
f

tapply(x, f, mean)

tapply(x, f, range)

```

# split #

`split` is like `tapply`, but without applying the summary statistics. It is useful in conjuction with `lapply` or `sapply`. `split` always returns a list

```{r, collapse=TRUE}
x <- c(rnorm(10), runif(10), rnorm(10,1))
f<- gl(3,10)
split(x, f)

## in conjuction with lapply

lapply(split(x,f), mean)

```

# Splitting a dataframe #

```{r, collapse=TRUE}
library(datasets)
head(airquality)

## calculate the mean within each month

s <- split(airquality, airquality$Month)
lapply(s, function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")]))


sapply(s, function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")]))

sapply(s, function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")], na.rm=TRUE))
```