---
title: "R Objects"
author: "Sarah Meister"
date: 2015-09-16
fontsize: 12pt
geometry: margin=1in
papersize: letterpaper
layout: page
sidebar: false
output: html_document
---
R is an object-oriented programing language, meaning everything we incounter in R is an object. 

R has several types of objects:


R has a 5 basic classes of objects:

* Character
* Numeric
* Integer
* Complex
* Logical

R has 5 different types of objects:

* Vectors
* Lists
* Matrices
* Factors
* Dataframes

# Vectors #

The most basic object is a vector. A vector is a sequence of objects of the same class. We can make vectors using the c() function

```{r}
a <- c(0.4, 1.2)         # numeric
b <- c(TRUE, FALSE)      # logical
c <- c("a","b","c")      # character
d <- c(1L, 2L)           # integer
e <- 1:10                # integer
f <- c(1+0i, 2+4i)       # complex
```

Empty vectors can be made using the vector() function

```{r}
x <- vector("numeric", length=10)
x
```

# Lists #

A list is a special type of vector that can contain objects of different classes. For instance, we can have a list with a numeric dataframe as one object within the list, and a string vector as another. 

```{r}
x <- list(1,"a", TRUE, 1+4i)
x
```

The elements of a list are seperated by double brackets. 
 
# Matrices #
Matrices are vectors with a dimension attribute. The dim attribute is an integer vector of length = 2 (nrow, ncol). 

```{r}
x <- matrix(1:6, nrow=2, ncol=3)

dim(x)

attributes(x)
```
 
Matrices are constructed column-wise, so entries start at the upper left corner and run down the columns. 

Matrices can be created directly from vectors by adding a dimension value

```{r}
x <- 1:10
x
dim(x) <- c(2,5)
x
```

We can also create matrices by column-binding or row-binding with the functions cbind() and rbind()

```{r}
x <- 1:3
y <- 10:12
cbind(x,y)
rbind (x, y)
```

# Factors #

Factors are used to represent catagorical data. They can be ordered or unordered.

* They are like an integer vector where each integer has a label. 
* Factors are more descriptive than integers e.g. "Male" and "Female" vs. 1 and 2
* Factors are treated specially by modelling functions like lm() and glm()

```{r}
x <- factor(c("yes", "yes", "no", "no", "yes"))
x
table (x)
unclass(x)
```

# Attributes #

R objects can have attributes, such as:

* names
* classes
* lengths
* dimensions (matrices/dataframes)
* user defined attributes, such as metadata

Attributes can be modified with the attributes() function
