---
title: "Introduction to R"
author: "Sarah Meister"
date: '2015-11-02'
output: html_vignette
layout: page
sidebar: no
published: false
---

# Starting R #

We will be using R studio for this course as it's a user-friendly GUI for R. You can also access R from Terminal (Mac) or Command Line (Windows) if you please. 

```{r, collapse=TRUE, eval=FALSE}
getwd()

setwd("/Users/thesarahmeister/")

dir.create("Desktop/practice-2015-10/sarah/intror")

setwd("Desktop/practice-2015-10/sarah/intror")

dir()

```

# Now let's code! #

Download the `airQuality.csv` file from our [Github repo](https://github.com/codeasmanuscript/practice-2015-10) and import it into R using the `read.csv` function


```{r, collapse=TRUE}
airQuality <- read.csv("airQuality.csv", header = TRUE)
```

Now let's look at the data using a few different functions

```{r, collapse=TRUE}
head(airQuality)

head(airQuality, n=8)

tail(airQuality)

str(airQuality)

summary(airQuality)

nrow(airQuality)

ncol(airQuality)

range(airQuality$Temp)

```

Lets alter our dataset a bit

```{r, collapse=TRUE}

names(airQuality$Solar.R) <- "Solar"

airQuality$Random <- rnorm(153, 1, 2)

```

Okay time to subset our data

```{r, collapse=TRUE}

rowThree <- airQuality[3, ]

colFour <- airQuality[ ,4]

ozone <- airQuality$Ozone

highTemp <- airQuality[airQuality$Temp > 80, ]

may <- airQuality[airQuality$Month == "May", ]

highTempOrMay <- airQuality[airQuality$Month == "May" | airQuality$Temp > 80, ]

lowTempAndSep <- subset(airQuality, airQuality$Month == "Sep" & airQuality$Temp < 70)

```

okay let's perform some basic data manipulations on our dataframe and subsetted data

```{r, collapse=TRUE, eval=FALSE}

mean(airQuality$Ozone) #returns an NA

mean(airQuality$Ozone, na.rm=TRUE)

colMeans(airQuality, na.rm=TRUE) #this one doesn't work

airQualNoMonth <- airQuality[,-5]

colMeans(airQualNoMonth, na.rm=TRUE)

sd(colFour)

sd(airQualNoMonth)
```

What happens if we have multiple columns that are non-numeric and we don't want to remove them all to use the built-in `colMeans()` function? We can make our **own** function!

```{r, collapse=TRUE}
columnmean <- function (y, remove.NA = TRUE) {
  nc <- ncol(y)
  means <- vector("numeric", length=0)
  for (i in 1:nc) {
    means[i] <- mean(y[,i], na.rm = remove.NA)
  }
  means
}

columnmean(airQuality)
```

Cool right? Now let's try a function with an if/else expression. 

```{r, collapse=TRUE}

above80 <- function(y) {
    y <- na.omit(y)
    nob <- length(y)
    high <- vector("numeric", length=0)
    low <- vector("numeric", length=0)
  for (i in 1:nob) {
    if (y[i] > 80) {
      high <- append(high, y[i])
    }
    else {
      low <- append(low, y[i])
    }
  }
  list(high,low)
}

above80(airQuality$Temp)

highlow <- above80(airQuality$Temp)

high <- highlow[[1]]

low <- highlow[[2]]
```

# Loop Functions #

The last thing for today are the very useful built-in loop functions. These functions have the word `apply` in them.

- `lapply`: loop over a list and evaluate a function on each element 
- `sapply`: same as `lapply` but it tries to simplify the result
- `apply`: apply a function over the margins of an array 
- `tapply`: apply a function over subsets of a vector (`lapply()` used with `split()` does the same thing)
- `mapply`: a multivariate version of `lapply` 

# lapply() #

Remember the function we made? It can be replaced with an `lapply()` loop.
```{r, collapse=TRUE}
lapply(airQuality, mean, na.rm = TRUE)
```

# sapply() #

Alternatively we can try `sapply()`. This will return a named numeric vector (simpified from a list)
```{r, collapse=TRUE}
sapply(airQuality, mean, na.rm = TRUE)
```

# apply() #

`apply()` first coerces your dataframe to a matrix, which means all the columns must have the same type. Because our dataframe does not meet this requirement, I'll be taking a subset of the data. However normally this function would be used with matrices.

```{r, collapse=TRUE}
apply(airQuality[,c(-5)], 2, sd, na.rm = TRUE)

apply(airQuality[,c(-5)], 1, sd, na.rm = TRUE)
```

# mapply #

`mapply` is a multivariate apply (like `lapply`) that applies a function over a set of arguments.
```{r, collapse=TRUE}
mapply(range, airQuality[,c(1,3,6)], na.rm = TRUE)
```

# tapply #

`tapply` is used to apply a function over subsets of a vector. It takes two arguements: a vector to apply to function on and a factor variable that subsets the vector. It will simplify the result, like sapply. 

```{r, collapse=TRUE}
tapply(airQuality$Ozone, airQuality$Month, range, na.rm = TRUE)
```

