---
title: "Introduction to R"
author: "Sarah Meister"
date: '2015-11-02'
output: html_vignette
fontsize: 8pt
header-includes: \input{../slideOptions.tex}
layout: page
classoption: xcolor=dvipsnames
sidebar: no
slide-level: 1
tag:
- Lessons
- Slides
- IntroR
categories:
- Lessons
- IntroR
---

# Starting R #

We will be using R studio for this course as it's a user-friendly GUI for R. You can also access R from Terminal (Mac) or Command Line (Windows) if you please. 

```{r, collapse=TRUE}
getwd()

setwd("/Users/thesarahmeister/")

setwd("Github/workshops/lessons/rintro")

```

# Now let's code! #

Download the airQuality.csv file from our [Github repo](https://github.com/codeasmanuscript/workshops/tree/master/lessons/rintro) and import it into R using the `read.csv` function


```{r, collapse=TRUE}
airQuality <- read.csv("airQuality.csv", header = TRUE)
```

Now let's look at the data using a few different functions

```{r, collapse=TRUE}
head(airQuality)

head(airQuality, n=8)

tail(airQuality)

str(airQuality)

summary(airQuality)

nrow(airQuality)

ncol(airQuality)

range(airQuality$Temp)

```

Lets alter our dataset a bit

```{r, collapse=TRUE}

names(airQuality$Solar.R) <- "Solar"

airQuality$Random <- rnorm(153, 1, 2)

```

Okay time to subset our data

```{r, collapse=TRUE}

rowThree <- airQuality[3, ]

colFour <- airQuality[ ,4]

ozone <- airQuality$Ozone

highTemp <- airQuality[airQuality$Temp > 80, ]

may <- airQuality[airQuality$Month == "May", ]

highTempOrMay <- airQuality[c(airQuality$Month == "May" | airQuality$Temp > 80), ]

lowTempAndSep <- subset(airQuality, airQuality$Month == "Sep" & airQuality$Temp < 70)

```

okay let's perform some basic data manipulations on our dataframe and subsetted data

```{r, collapse=TRUE}

mean(airQuality$Ozone)

mean(airQuality$Ozone, na.rm=TRUE)

colMeans(airQuality, na.rm=TRUE)

airQualNoMonth <- airQuality[,-5]

colMeans(airQualNoMonth, na.rm=TRUE)

sd(colFour)

sd(airQualNoMonth)
```


```{r, collapse=TRUE}
columnmean <- function (y, remove.NA = TRUE) {
  nc <- ncol(y)
  means <- numeric(nc)
  for (i in 1:nc) {
    means[i] <- mean(y[,i], na.rm = remove.NA)
  }
  means
}
```

```{r, collapse=TRUE}

```

```{r, collapse=TRUE}

```