---
title: "Introduction: Data wrangling in R"
author:
    - Luke W. Johnston
date: 2015-10
layout: page
sidebar: false
tag:
    - Lessons
    - Introduction
    - Wrangling
    - R
categories:
    - Lessons
    - Wrangling
    - R
permalink: lessons/topic/
---

* Introductions
* State goal for workshop
* Item 2

# Learning objectives: #

1. Objective one
2. Objective two

# Import/export your data #

You'll need to import your data into R to analyze it.  I'm assuming
the data is already cleaned and ready for analysis.  If at any time
you need help with a command, use the `?` command, appended with the
command of interest (eg. `?write.csv`).  R comes with many internal
datasets that you can practice on.  The one I'm going to use is the
`swiss` dataset.

```{r importData}

## Export
write.csv(swiss, file = 'swiss.csv') # Export
## Import
ds <- read.csv('swiss.csv') # Import

```

# Viewing your data #

R has several very useful and easy tools for quickly viewing your
data.  `head()` shows the first few rows of a data.frame (a structure
for storing data that can include numbers, integers, factors, strings,
etc).  `names()` shows the column names.  `str()` shows the structure,
such as what the object is, and its contents.  `summary()` shows a
quick description of the summary statistics (means, median, frequency)
for each of your columns.  `class()` is like `str()` but only shows
the top level name of the object, so eg. while a data.frame contains
multiple columns that `str()` would show, `class()` would only show
that the object is a "data.frame".

```{r viewData}
head(ds)
names(ds)
str(ds)
summary(ds)
```

# Wrangling your data #

Data wrangling is a bit tedious in base R.  So I'm using two packages
designed to make this easier.  Load packages by using the `library()`
function.  `dplyr` comes with a `%>%` pipe function (via the
`magrittr` package), which works similar to how the Bash shell `|`
pipe works.  The command on the right-hand side takes the output from
the command on the left-hand side, just like how a plumbing pipe works
for water.

The four lines of code below using `tbl_df` are all the same.  The `.`
object represents the output from the pipe, but it doesn't have to be
used as using `%>%` implies also using `.`.  `tbl_df` makes the object
into a `tbl` class, making printing of the output nicer.

```{r wrangle}

library(dplyr)
tbl_df(ds)
ds %>% tbl_df()
ds %>% tbl_df
ds %>% tbl_df(.)

## Let's put it into a new object
ds2 <- tbl_df(ds)

```

Again, these next lines are the same. `select` does as it says: select
the column from the dataset.

```{r wrangleSelect}

select(ds2, Education, Catholic, Fertility)
ds2 %>% select(Education, Catholic, Fertility)
ds2 %>% select(., Education, Catholic, Fertility)

```

You can rename columns either using `rename` or `select` (the new name
is on the left hand side, so `newname = oldname`).  However, with the
`select` command, only that column gets selected, while `rename`
selects all columns.

```{r wrangleRename}
ds2 %>% rename(County = X)
ds2 %>% select(County = X)
```

You can subset the dataset using `filter`.  Note the double equal sign
`==` for testing if 'Examination' is equal to 15.  A single `=` is
used for something else (assigning things to objects).

```{r wrangleFilter}

filter(ds2, Catholic < 20, Examination == 15)
ds2 %>% filter(Catholic < 20, Examination == 15)
ds2 %>% filter(., Catholic < 20, Examination == 15)
## For string/factor variables
ds2 %>% filter(X == 'Aigle')

```

We can start chaining these commands together using the `%>%` command.
There is no limit to how long a chain can be.  `arrange`
sorts/orders/re-arranges the column Education in ascending
order. `mutate` creates a new column.

```{r wrangleChain}

ds2 %>%
  filter(Catholic > 20) %>%
  select(Education, Fertility) 
  
ds2 %>%
  filter(Catholic > 20) %>%
  select(County = X, Education, Fertility, Agriculture) %>%
  arrange(Education) %>%
  mutate(infertile = ifelse(Fertility < 50, 'yes', 'no'),
         testing = 'Yes' ## Create a testing column to show how mutate works.
         )

```

To get the data into a nicer and more analyable format, you can use
the `tidyr` package.  See what `gather` does in the code below.  Then
see what `spread` does.  Note that you can remove a column by having a
minus `-` sign in front of a variable when you use `select`.

```{r reorg}

library(tidyr)
## Compare this:
ds2 %>%
  select(-Infant.Mortality) %>%
  rename(County = X)

## With this:
ds2 %>%
  select(-Infant.Mortality) %>%
  rename(County = X) %>%
  gather(Measure, Value, -County)

## And back again:
ds2 %>%
  select(-Infant.Mortality) %>%
  rename(County = X) %>%
  gather(Measure, Value, -County) %>%
  spread(Measure, Value)

```

Combined with `dplyr`'s `group_by` and `summarise` you can quickly
summarise data or do further, more complicated analyses. `group_by`
makes it so further analyses or operations work on the groups.
`summarise` transforms the data to only contain the new variable(s)
created, in this case the mean.

```{r reorgChain}

ds2 %>%
  select(-X) %>%
  gather(Measure, Value) %>%
  group_by(Measure) %>%
  summarise(mean = mean(Value))

```

# Materials for this lesson: #

* [Slides](slides/)
* [Cheatsheet](cheatsheet/)
* [Assignment](assignment/)

Other resources can be found [here](../resources/).

